### 2021-04-28 latex/oscillations.tex
update

\documentclass[twocolumn]{article}
\input{commands.tex}
\usepackage{physics,amsmath, amsfonts, siunitx, amssymb,cite}
\usepackage[utf8]{inputenc}
\usepackage{xr-hyper}
%\usepackage{hyperref}
\externaldocument{SM}

\begin{document}
\section{Neutrino Masses and Oscillations}\label{ch:oscillation}
As we saw in Eq.~\ref{eq:j_CC}
, the neutrino fields $\nu_\alpha$ only couple to the associated lepton fields $\ell_\alpha$, conserving the lepton number $L_\alpha$. We will now introduce two separate extensions to this part of the Standard Model.

@@

Substituting the new transformation from Eq.~\ref{eq:nu_rotation} into the weak charged current, we get
\begin{align}\label{eq:j_CC2}
    j^\rho_L &= 2\bar{\nu}^\prime_{\alpha L} \gamma^\rho \ell_{\alpha L}^\prime \nonumber \\
             &= 2\bar{\nu}_{k L} V^{\prime \nu \dagger}_{k \alpha}V^{\prime \ell}_{\alpha \alpha} \gamma^\rho  \ell_{\alpha L}
\end{align}
Call $V^{\prime \nu \dagger}_{k \alpha}V^{\prime \ell}_{\alpha \alpha} = U^\dagger_{k \alpha}$
Now, the current in Eq.~\ref{eq:j_CC2} conserves lepton number, since the neutrino field with flavor $\alpha$ only couples to the lepton field with flavor $\alpha$. However, the Higgs-lepton Yukawa Lagrangian 
in Eq.~\ref{eq:L_H} violates lepton number conservation since it couples the charged lepton flavor $\alpha$ to the neutrino mass eigenstate $k$, which is a superposition of flavors. There is no transformation that leaves both the interaction and kinetic Lagrangian invariant.
We now have 
\begin{align}
    j^\rho_L &= 2 U^*_{\alpha k} \bar{\nu}_{k L} \gamma^\rho  \ell_{\alpha L}
\end{align}

Derivation of neutrino oscillation formula
%\bibliographystyle{apalike}
%\bibliography{ref}
\end{document}

>>>>

\documentclass[draft=True]{revtex4-2}
\input{commands.tex}
\usepackage{physics,amsmath, amsfonts, siunitx, amssymb, graphicx, slashed}
\usepackage[utf8]{inputenc}
%\usepackage{hyperref}
\usepackage{xr-hyper}
\externaldocument{SM}

\begin{document}

\section{Neutrino Masses and Oscillations}\label{ch:oscillation}
As we saw in Eq.~\ref{eq:j_CC}
, the neutrino fields $\nu_\alpha$ only couple to the associated lepton fields $\ell_\alpha$, conserving the lepton number $L_\alpha$. We will now introduce two separate extensions to this part of the Standard Model.

@@

Substituting the new transformation from Eq.~\ref{eq:nu_rotation} into the weak charged current, we get
\begin{align}\label{eq:j_CC2}
    j^\rho_L &= 2\bar{\nu}^\prime_{\alpha L} \gamma^\rho \ell_{\alpha L}^\prime \nonumber \\
             &= 2\bar{\nu}_{k L} V^{\prime \nu \dagger}_{k \alpha}V^{\prime \ell}_{\alpha \alpha} \gamma^\rho  \ell_{\alpha L}
\end{align}
Call $V^{\prime \nu \dagger}_{k \alpha}V^{\prime \ell}_{\alpha \alpha} = U^\dagger_{k \alpha}$
Now, the current in Eq.~\ref{eq:j_CC2} conserves lepton number, since the neutrino field with flavor $\alpha$ only couples to the lepton field with flavor $\alpha$. However, the Higgs-lepton Yukawa Lagrangian 
in Eq.~\ref{eq:L_H} violates lepton number conservation since it couples the charged lepton flavor $\alpha$ to the neutrino mass eigenstate $k$, which is a superposition of flavors. There is no transformation that leaves both the interaction and kinetic Lagrangian invariant.
We now have 
\begin{align}\label{eq:j_CC3} %TODO: put sums on all above?
    j^\rho_L &= 2 \sum_\alpha \sum_k U^*_{\alpha k} \bar{\nu}_{k L} \gamma^\rho  \ell_{\alpha L}
\end{align}
Now, the Fourier expansion of the field operator $\bar{\nu}_{k L}$ in Eq.~\ref{eq:j_CC3} contains creation operators $a^\dagger_{\nu_k}$
of massive neutrinos with mass $m_k$. This means that the summation over the mass index $k$ constructs a flavor neutrino, which interacts with the charged lepton field $\ell_{\a L}$.
In other words, the charged current generates a flavor neutrino $\na$, which is a superposition of the mass eigenstates 
$\nu_k$ with weights $U_{\a k}^*$. In the ket-formalism, we express this as
\begin{align}\label{eq:osc_1}
    \ket{\na} = \sum_k U^*_{\a k} \ket{\nu_k}\,.
\end{align}
It is the mass eigenstates $\ket{\nu_k}$ that are eigenstates of the Hamiltonian, with eigenvalues
\begin{align}\label{eq:disp}
    E_k = \sqrt{\vec{p}^2 + m_k^2}\,.
\end{align}
The solution to the time-dependent Schrödinger equation 
\begin{align}
    i \dv{}{t} \ket{\nu_k(t)} = H\ket{\nu_k(t)}\,,
\end{align}
gives us a plane wave evolution to the mass states as 
\begin{align}
    \ket{\nu_k(t)} = e^{-iE_kt}\ket{\nu_k}\,.
\end{align}
Inserting the plane wave solution into Eq.~\ref{eq:osc_1}, we get 
\begin{align}\label{eq:nat}
    \ket{\na(t)} = \sum_k U^*_{\a k} e^{-iE_kt}\ket{\nu_k}\,.
\end{align}
Now we know how to evolve and combine the mass eigenstates to form a flavor eigenstate, but how about the reverse? We swap the index $k\to j$ in Eq.~\ref{eq:osc_1} and multiply by $U_{\a k}$:
\begin{align}\label{eq:osc_k}
    \sum_\a U_{\a k}\ket{\na} &= \sum_{\alpha, j} U_{\a k} U_{\a j}^* \ket{\nu_j} \nonumber \\
    \sum_\a U_{\a k}\ket{\na} &= \sum_{j} \delta_{kj} \ket{\nu_j} \nonumber \\
    \sum_\a U_{\a k}\ket{\na} &= \ket{\nu_k}\,,
\end{align}
where we have used the unitarity of the leptonic mixing matrix. Eqs.~\ref{eq:osc_1} and \ref{eq:nat} yield 
\begin{align}
    \ket{\na(t)} &= \sum_k U^*_{\a k} e^{-iE_kt}\ket{\nu_k} \nonumber \\
    \ket{\na(t)} &= \sum_k U^*_{\a k} e^{-iE_kt}\left(\sum_\b U_{\b k}\ket{\nb}\right) \nonumber \\
    \ket{\na(t)} &= \sum_{k,\b} U^*_{\a k}U_{\b k} e^{-iE_kt} \ket{\nb}\,.
\end{align}
The probability of the flavor transition $\na \to \nb$ at time $t$ is $\abs{\bra{\nb}\ket{\na (t)}}^2$:
\begin{align}
    P_{\na \to \nb}(t) = \sum_{k,j} U^*_{\a k}U_{\b k}U^*_{\b j}U_{\a j} e^{-i(E_k-E_j)t}\,.
\end{align}

We assume the neutrino masses $m_k$ to be extremely small compared to their associated energies $E_k$. Thus, $v\approx 1$, and $\abs{\vec{p}} \approx E$ making the energy-dispersion relation of Eq.~\ref{eq:disp} to first order:
\begin{align}
    E_k &= \sqrt{\vec{p}^2 + m_k^2} \nonumber \\
        &= \vec{p}^2\sqrt{1 + \frac{m_k^2}{\vec{p}^2}} \nonumber \\
        &\approx E + \frac{m_k^2}{2E}
\end{align}
Hence, the exponential can be simplified, and simplifying the notation $P_{\na \to \nb}(t) \to P_{\ab}(t)$ we get 
\begin{align}
    P_{\ab}(t) = \sum_{k,j} U^*_{\a k}U_{\b k}U^*_{\b j}U_{\a j} e^{-i(m_k^2-m_j^2)t/2E}\,.
\end{align}

Now, our approximation $v\approx 1$ implies $x\approx t $, thus
\begin{align}
    P_{\ab}(x) &= \sum_{k,j} U^*_{\a k}U_{\b k}U^*_{\b j}U_{\a j} e^{-i(m_k^2-m_j^2)x/2E} \nonumber \\
                       &= \sum_{k,j} U^*_{\a k}U_{\b k}U^*_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E})\,,
\end{align}
where we in the last step have defined the \emph{mass squared-difference} $\dm_{kj} = m_k^2 - m_j^2$. Thus, it is impossible
to measure the mass states $m_k$ through neutrino oscillations, since they depend on the difference between the two squared mass states instead of the
masses themselves. 
Squaring the unitarity condition $\sum_k U_{\a k}U_{\b k}^* = \delta_{\a \b}$ yields 
\begin{align}
    \sum_k \abs{U_{\a k}}^2\abs{U_{\b k}}^2 = \delta_{\a \b} - 2 \sum_{k>j}\Re[U_{\a k}^* U_{\b k} U_{\a j} U_{\b j}^*]
\end{align}
In this work, we will always assume the mixing matrix $U$ to be real. 
We thus write 
\begin{align}\label{eq:Pab}
    P_{\ab} &= \sum_{k,j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &=  \sum_k \abs{U_{\a k}}^2\abs{U_{\b k}}^2 + \nonumber \\
            &\hspace{2em}+ \sum_{k \neq j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2 \sum_{k>j}U_{\a k} U_{\b k} U_{\b j} U_{\a j} \nonumber \\
            & \hspace{3em}+\sum_{k \neq j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2 \sum_{k>j}U_{\a k} U_{\b k} U_{\b j} U_{\a j} \nonumber \\
            &\hspace{3em}+ 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \left[1-\exp(-i\frac{\dm_{kj}x}{2E})\right] \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \left[1-\cos(\frac{\dm_{kj}x}{2E})\right] \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \sin^2{\left(\frac{\dm_{kj}x}{4E}\right)}\,,
\end{align}
which is the probability of neutrino vacuum oscillations. %TODO: maybe plot?
The calculation for antineutrinos yield the same result if one assumes the realness of the mixing matrix.

\subsection{Matter effects}
\begin{figure}\label{fig:w_and_z}
    \includegraphics[width=0.1\textwidth]{figures/w-boson.pdf}
    \hspace{2cm}
    \includegraphics[width=0.1\textwidth]{figures/z-boson.pdf}
\end{figure}
In this work, we are concerned about the interactions with the neutrino and Earth-like matter, i.e. electrons, protons, and neutrons. 
The possible interactions are shown in Fig. ~\ref{fig:w_and_z}. The left panel shows that the only flavor that can go through charged current (CC) 
interactions is the electron flavor. This is because the Earth doesn't consist of any muons or tau particles. 
The right panel shows any neutrino flavor interaction via the neutral current (NC) with Earth-like matter, mediated by the neutral $Z$ boson
 we see weak forward CC and NC scattering. 
The interaction mediated by the $W$ boson will give rise to a effective matter potential $V_{CC}$, while
the $Z$ boson is responsible for $V_{NC}$. Our task is now to find expressions for these.

\subsection{Effective matter potentials}

We start with the effective Hamiltonian for the CC process. The Feynman rules for the left panel give us 
\begin{align}
    H = \frac{G_F}{\sqrt{2}}\left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, e \right] \left[\bar e \,\gamma_\rho\, (1- \gamma^5)\, \ne \right]
\end{align}
By using the Fierz transformation 
\begin{align}
    \mathcal{L}^{V-A}(\psi_1,\psi_2,\psi_3,\psi_4) = \mathcal{L}^{V-A}(\psi_1,\psi_4,\psi_3,\psi_2)\,,
\end{align}
we can permute the terms inside the brackets, yielding
\begin{align}\label{eq:H_fierz}
    H = \frac{G_F}{\sqrt{2}}\left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \right] \left[\bar e \,\gamma_\rho\, (1- \gamma^5)\, e \right]\,.
\end{align}
Now, lets consider a finite volume $V$ with electron states defined as 
\begin{align}\label{eq:e_states}
    \ket{e(p_e,h_e)} = \frac{1}{2E_eV}a_e^{(h_e)\dagger}(p_e)\ket{0}\,,
\end{align}
i.e. using the creation operator $a_e^{(h_e)\dagger}(p_e)$ to create electron states from vacuum with momenta $p_e$, energy $E_e$, and helicity $h_e$.
The density distribution of electrons in $V$ is $f(E_e,T)$, which we normalize to the total number of electrons as we integrate out the momenta $p_e$:
\begin{align}\label{eq:e_density}
    \int \dd p_e^3 f(E_e,T) = N_e V = n_e
\end{align}
Here, the electron density $N_e$ will ultimately determine the strength of the effective matter potential. 
To obtain the average effective Hamiltonian, project it on the electron states in Eq.~\ref{eq:e_states} and integrate over the density and sum over the helicities:

\begin{align}\label{eq:avg_H1}
    \bar H &= \int \dd p_e^3 \bra{e(p_e,h_e)} \times \frac{1}{2} \sum_{h_e} H   f(E_e,T) \ket{e(p_e,h_e)} \nonumber \\
           &= \frac{G_F}{\sqrt{2}} \int \dd p_e^3 \bra{e(p_e,h_e)} \left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \right]  f(E_e,T) \times \frac{1}{2} \sum_{h_e} \left[\bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x) \right] \ket{e(p_e,h_e)} \nonumber \\
           &= \frac{G_F}{\sqrt{2}} \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \int \dd p_e^3  f(E_e,T) \times \frac{1}{2} \sum_{h_e} \bra{e(p_e,h_e)} \bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x)   \ket{e(p_e,h_e)}\,.
\end{align}
First, calculate the sum using trace technology %TODO: explain this more
\begin{align}\label{eq:helicity_sum}
    \frac{1}{2} \sum_{h_e} \bra{e(p_e,h_e)} \bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x)   \ket{e(p_e,h_e)} &= \frac{1}{4E_e V} \sum_{h_e} \bar{u}_e^{h_e}(p_e) \,\gamma_\rho\, (1- \gamma^5)\, u_e^{h_e}(p_e) \nonumber \\
    &= \frac{1}{4 E_e V} \Trace{\left[ \sum_{h_e} \bar{u}_e^{h_e}(p_e) u_e^{h_e}(p_e) \,\gamma_\rho\, (1- \gamma^5)\, \right]} \nonumber \\
    &= \frac{1}{4 E_e V} \Trace{\left[ (\slashed{p}_e + m_e ) \,\gamma_\rho\, (1- \gamma^5)\, \right]} \nonumber \\
    &= \frac{(p_e)_\rho}{E_e V}
\end{align}
Eq.~\ref{eq:avg_H1} now becomes 
\begin{align}\label{eq:avg_H2}
    \avg{H} &= \frac{G_F}{\sqrt{2}E_e V} \bar\ne \, (1- \gamma^5)\, \ne \int \dd p_e^3\, \slashed{p}_e\, f(E_e,T)\,.
\end{align}
Expand the integral, and use the fact that $\vec{p}_e$ is odd:
\begin{align}
    \int \dd p_e^3\, \slashed{p}_e\, f(E_e,T) &= \int \dd p_e^3\, f(E_e,T) (\gamma^0 E_e - \vec{p}_e \cdot \vec{gamma}) \nonumber \\
                                              &= \int \dd p_e^3\, f(E_e,T) \gamma^0 E_e \nonumber \\
                                              &= \gamma_0 E_e N_e V\,.
\end{align}
Inserting this into Eq.~\ref{eq:avg_H2}, we have
\begin{align}
    \avg{H} &= \frac{G_F N_e}{\sqrt{2}} \bar{\nu}_e \, (1- \gamma^5)\, \ne \gamma_0 \nonumber \\
            &= \sqrt{2}G_F N_e \bar{\nu}_{Le} \gamma^0 \nu_{Le}\,,
\end{align}
where the projection operator $(1- \gamma^5)$ ensures that only the left-hand compontent of the neutrino fields interact weakly. Thus 
\begin{align}\label{V_CC1}
    V_{CC} = \sqrt{2}G_F N_e
\end{align}

For neutral current, we replace the electron field $e(x)$ in Eq.~\ref{eq:H_fierz} by the fermion field $f(x)$, and the projection operator 
$(1-\gamma^5)$ with $(g_V^f - g_A^f\gamma^5)$. Again, the $\gamma^5$ will cause the spacial component of $p_f$ to disappear after integration, and the 
only difference between the average effective Hamiltonian for the neutral current is then the factor $g_V^f$:
\begin{align}
    V^f_{NC} = \sqrt{2}G_F N_A g_V^f\,.
\end{align}
Summing over the fermions, and assuming electrical neutrality and equal abundance of protons and neutrons, we have
\begin{align}
    V_{NC} &= \sum_{f \in {e,p,n}} V^f_{NC} \nonumber \\
           &= \sqrt{2}G_F N_A \sum_{f \in {e,p,n}} g_V^f \nonumber \\
           &= \sqrt{2}G_F N_A\left[ -\frac{1}{2}+2\sin^2{(\theta_W)} + \frac{1}{2}-2\sin^2{(\theta_W)} -\frac{1}{2} \right] \nonumber \\
           &= -\frac{1}  {\sqrt{2}} G_F N_e\,,
\end{align}
where the electrical neutrality condition allows us to simply sum the vectorial couplings together, cancelling the electron and proton contributions (and hence, also the mixing angle dependence).
\subsection{Matter oscillations}
Since only $\ne$ undergo CC interactions, the $V_{CC}$ potential is zero for all other flavors. However, since all flavors, undergo NC interactions the total matter potential in matrix form is:
\begin{align}
    V = \begin{bmatrix}
        V_{CC} + V_{NC} & 0 & 0 \\
        0 & V_{NC} & 0 \\
        0 & 0 & V_{NC} \\
    \end{bmatrix}
\end{align}
%\bibliographystyle{apalike}
%\bibliography{ref}
\end{document}
